---
title: "Unemployment in STEM"
subtitle: "STAT 650 Advanced R Final Project"
author: 
  - name: "Lydia Gibson"
  - name: "Ken Vu"
editor: source
format:
  pdf:
    echo: false
    documentclass: scrartcl
    classoption: ["oneside", "open=any"]
    number-sections: false
    template-partials:
      - "before-body.tex"
      - "_titlepage.tex" 
    include-in-header: 
      - "in-header.tex"
    toc: true
---

\newpage
  
## Introduction

Does obtaining a STEM (Science, Technology, Engineering and Mathematics) degree truly ensure that you will obtain a lucrative career? While we know that all STEM degrees do not secure the same level of earnings, surely they are at least equal in employment opportunities. Unfortunately, this is not the case as we've observed.

Although government agencies and politicians have long espoused the importance of our youth pursuing careers in STEM, the availability of these careers does not always match the market demand of our ever growing workforce. Although entry into some STEM careers is possible with a HS diploma or associates degree, often a minimum of a 4-years bachelor degree is mandatory and for others, a graduate degree such as a masters or doctoral degree is necessary for entry into specific careers.

As projected, many STEM careers which require degrees do have faster than average employment growth, but this is not true across the board for all STEM careers requiring degrees. In those instances where the growth of the workforce outpaces that of the growth of employment opportunities, we end up with highly educated individuals who are either unemployed or underemployed. 

## Data Description


\setlength\parindent{24pt}        

The data was obtained from the American Community Survey 2010-2012 Public Use Microdata Series and used in the fivethirtyeight story [The Economic Guide To Picking A College Major](https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/). Our data dimensions are **76 rows** by **12 columns**. The variables are: `Rank`, `Major_code`, `Major`, `Major_category`, `Total`, `Men`, `Women`, `ShareWomen`, `Employed`, `Unemployed`, `Unemployment_rate`, and `Median`. Below are descriptions of each variable.

- `Median`: Median earnings of full-time, year-round workers
- `Rank`: Rank by median earnings
- `Major_code`: Major code, FO1DP in ACS PUMS
- `Major`: Major description
- `Major_category`: Category of major from Carnevale et al
- `Total`: Total number of people with major
- `Men`: Male graduates
- `Women`: Female graduates
- `ShareWomen`: Women as share of Total
- `Employed`: Number employed (ESR == 1 or 2)
- `Unemployed`: Number unemployed (ESR == 3)
- `Unemployment_rate`: Unemployed / (Unemployed + Employed)



## Results

```{r include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
```

```{r include=FALSE}
#Getting in women in stem data
women_stem <- read_csv("women-stem.csv")
#women_stem
```


```{r include=FALSE}
#Getting in `recent_grad` data set
recent_grads <- read_csv("recent-grads.csv")
#recent_grads
```



```{r}
#Merging the two data sets by major
full_data <- women_stem %>%
  inner_join(recent_grads %>% 
               select(Major, contains("employ")), by="Major")

```

```{r, warning=F, fig.height = 8, fig.width = 8}
# Get the outliers
outlier_pts <- full_data %>%filter(Median > 100000 |(Median > 60000 & Major_category == "Physical Sciences"))

full_data %>%
  ggplot(aes(x=Major_category,y=Median,
             color=Major_category,
             size=Unemployment_rate)) +
  geom_jitter(alpha=1/4) + # set transparency of points
  # label outliers
  geom_text(data=outlier_pts, aes(label=str_wrap(Major,width=15), size=0.01), hjust=0.3, vjust=1.25, color="Black") + 
  labs(x="Major Category", y="Median Salary ($ Thousands)",
       title="Median Salary by STEM Major Category",
       size="Unemployment Rate") +
  theme(legend.position ="top",
        plot.title=element_text(hjust=0.5)) + 
  scale_y_continuous(breaks=c(40000,60000,80000,100000),
                     labels=c("40","60","80","100")) +
  guides(color=F, # remove legend for color
         size=guide_legend( # change size legend
           override.aes = list(alpha = 1#,
                               #size = c(3,4.5,5.8,7)
                               )
           )
         ) +
  # Wrap text for major categories
  scale_x_discrete(labels = scales::label_wrap(15)) +
  scale_size_binned(breaks = c(0.04, 0.06, 0.08, 0.10, 0.12))

  
```


```{r}
#get a data frame with summary statistics for all major categories
dat_stats_all <- recent_grads %>%
  group_by(Major_category) %>%
  summarise(UnemploymentRate = sum(Unemployed)/sum(Employed),
            AvgMedianSalary = mean(Median),
            Workforce = sum(Unemployed) + sum(Employed)) %>%
  mutate(Major_category = Major_category %>% as.factor() %>%
           fct_reorder(UnemploymentRate),
         STEM_Major = ifelse(!(Major_category %in% unique(women_stem$Major_category)),
                             "No","Yes") %>% as.factor() %>%
           fct_relevel("Yes"))
#dat_stats_all
```


```{r fig.height = 8, fig.width = 10}
avg_unemploy_rate = mean(recent_grads$Unemployment_rate)
dat_stats_all %>%
  ggplot(aes(y=Major_category, 
             x=UnemploymentRate, fill=STEM_Major)) +
  geom_col() + 
  #wrap y-axis label text
  scale_y_discrete(labels = scales::label_wrap(20)) +
  # add breaks to x-axis in 0.01 intervals
    scale_x_continuous(
    breaks = c(0.0, 0.01, 0.02, 0.03, 0.04, 0.05, 0.06, 0.07, 0.08, 0.09, 0.10, 0.11)) +
  # manually change fill colors
  scale_fill_manual(values=c("#0066CC","#bebebe")) +
  # add red line representing average unemployment rate 
  geom_vline(xintercept=avg_unemploy_rate,
             color="red", linetype=2) + 
  labs(x="Unemployment Rate", y="Major Category",
       title="Unemployment Rate by Major Category",
       fill="STEM") + 
  # change title size
  theme(plot.title=element_text(hjust=0.5,size=16)) + 
  # annotate vertical line
  annotate("text",x=avg_unemploy_rate, 
         y=1, label=paste("avg=", "","", "","", "","",round(avg_unemploy_rate,4)), color="red", size=4)
  
```


```{r}
#table for each major category with the unemployment rate and the average median salary
df1 <- dat_stats_all %>% 
  arrange(desc(AvgMedianSalary)) %>%
  select(c("Major_category","AvgMedianSalary","UnemploymentRate"))
names(df1) <- c("Major Category", "Average Median Salary ($)",
                "Unemployment Rate")
df1 %>% 
  kable() %>% 
  # add borders everywhere
  kable_styling(bootstrap_options = "bordered",
                full_width = FALSE) %>%
  # Add borders left and right as well as wrap text
  column_spec(c(1), border_left=T, width="7em") %>%
  column_spec(c(3), border_right=T, width="8em") %>%
  # wrap text for 2nd column
  column_spec(c(2), width="10em") %>%
  # make column headers bold
  row_spec(c(0), bold=T, font_size = 12)
```

## Conclusion

In conclusion, we see that the association between earnings and unemployment is indeed weak as evidenced by the relatively low earnings with the major category of health care yet lower than average unemployment rate in contrast to the higher earnings in the engineering major category yet higher than average unemployment. Also, we found that those earning degrees in STEM major categories of Computers & Mathematics, Biology & Life Sciences, and Engineering had a higher than average unemployment level across all STEM and non-STEM major categories. 

Therefore, we cannot conclude that obtaining a degree in a STEM field is the magic bullet to ensuring lucrative employment. This may be due to the fact that the varying occupations requiring such degrees have differing levels of growth and employment levels. Ideally, potential workers will want to aim for fields with many job openings relative to their employment size, such as those pertaining to healthcare or computer-related careers. Other occupations (particularly, certain engineering or mathematics positions) may have small employment levels with lower than average expected growth.

Overall, many STEM careers do enjoy median income above the national average so they are still worth striving for compared to other non-STEM careers with high unemployment and lower median salary.

\newpage

## Bibliography

  Casselman, B. “The Economic Guide to Picking a College Major.” FiveThirtyEight, FiveThirtyEight, 12 Sept. 2014, https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/. 

Terrel, N. (2007). Stem occupations: High-tech jobs for a high-tech economy. Occupational Outlook Quarterly. Retrieved September 30, 2022, from https://www.bls.gov/careeroutlook/2007/spring/art04.pdf 

Vilorio, D. (2014). Stem 101: Intro to Tomorrow's Jobs - Bureau of Labor Statistics. Occupational Outlook Quarterly. Retrieved September 30, 2022, from https://www.bls.gov/careeroutlook/2014/spring/art01.pdf 

## Code Appendix

For supplementary R script, visit <https://github.com/lgibson7/Unemployment-in-STEM>
